---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 10 @ 11:59PM
author: Cyrene Arputhasamy (705527395)
format:
  html:
    theme: cosmo
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    cache: false    
    echo: true
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
---

Display machine information for reproducibility:
```{r}
#| eval: true
sessionInfo()
```

Load necessary libraries (you can add more as needed).
```{r setup}
library(data.table)
library(lubridate)
library(R.utils)
library(tidyverse)
```

MIMIC data location
```{r}
mimic_path <- "~/mimic"
```

In this exercise, we use tidyverse (ggplot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2023winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

Display the contents of MIMIC data folder. 
```{r}
system(str_c("ls -l ", mimic_path, "/"), intern = TRUE)
system(str_c("ls -l ", mimic_path, "/core"), intern = TRUE)
system(str_c("ls -l ", mimic_path, "/hosp"), intern = TRUE)
system(str_c("ls -l ", mimic_path, "/icu"), intern = TRUE)
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. 

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)
```{r}
getwd()
setwd("~/mimic/core")

system.time(read.csv("admissions.csv.gz"))
system.time(read_csv("admissions.csv.gz", show_col_types = FALSE))
system.time(fread("admissions.csv.gz"))
```
It took 3.824 seconds to read the file with `read.csv`, 1.256 seconds to read
with `read_csv`, and 0.6 seconds to read with `fread`. The fasted method seems 
to be using `fread`. There is a difference in the default parsed data types. 

For later questions, we stick to the `read_csv` in tidyverse.

## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/icustays.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `icustatys.csv.gz` as a tibble `icustays_tble`. 
```{r}
setwd("~/mimic/icu")
icustays_tble <- as_tibble(read_csv("icustays.csv.gz", show_col_types = FALSE))
```

2. How many unique `subject_id`? Can a `subject_id` have multiple ICU stays? 
```{r}
length(unique(icustays_tble$subject_id))
length(icustays_tble$subject_id)
```
There are 53,150 unique entries for `subject_id`. It appears that a `subject_id`
can have multiple ICU stays since this number differs from the length of the 
entire column. According to the documentation, the `icustays` table is derived
from the `transfers` table, only including rows where the ward is an ICU. There 
are multiple idetifiers used in this table. `subject_id` is unique to the 
patient. `stay_id` is unique to the patient ward stay. This means a `subject_id`
may have multiple ICU stays due to transfers between wards, or different hospital
visits(`hadm_id`).

3. Summarize the number of ICU stays per `subject_id` by graphs. 
```{r}
stays_per_id <- icustays_tble %>% 
  count(subject_id) %>% 
  ggplot(aes(x=n)) + 
  geom_histogram(binwidth = 1, color="darkblue", fill="lightblue") 
stays_per_id
```

4. For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`. (Hint: `slice_min` and `slice_max` may take long. Think alternative ways to achieve the same function.)
```{r}
icustays_tble <- icustays_tble %>% group_by(subject_id) %>% filter(intime == min(intime))
#head(arrange(icustays_tble, subject_id, intime))
```

## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/hosp/admissions/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/core/admissions.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `admissions.csv.gz` as a tibble `admissions_tble`.
```{r}
setwd("~/mimic/core")
admissions_tble <- as_tibble(read_csv("admissions.csv.gz", show_col_types = FALSE))
```

2. Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadm_id`.
```{r}
admissions_tble <- admissions_tble %>%
  inner_join(icustays_tble, by = c("subject_id", "hadm_id"))
```

3. Summarize the following variables by graphics. 

    - admission year  
    - admission month  
    - admission month day  
    - admission week day  
    - admission hour (anything unusual?)  
    - admission minute (anything unusual?)  
    - length of hospital stay (anything unusual?)    
#### Admission Year
```{r}
admissions_tble$admit_year <- year(admissions_tble$admittime)
ggplot(admissions_tble, aes(admit_year)) + 
  geom_area(stat = "bin", fill = "lightblue", color = "darkblue") + 
  ggtitle("Number of Admits by Year") + 
  xlab("Year")
```
There is a wide range of years in the data set which are offset randomly per 
`subject_id` to de-identify patients. A single date shift was assigned to each 
`subject_id`. Thus the data for a single patient are consistent across modules, 
but distinct patients are not temporally comparable. 

#### Admission Month
```{r}
admissions_tble$admit_month <- month(admissions_tble$admittime) %>% 
  factor(levels = c(01, 02, 03, 04, 05, 06, 07, 08, 09, 10, 11, 12),
         labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep",
                    "Oct", "Nov", "Dec"))
ggplot(admissions_tble, aes(admit_month)) +
  geom_bar(fill = "lightblue", color = "darkblue")
```
There are slightly less admits in February than the other months, but they're
all generally around the same number. 

#### Admission Month Day
```{r}
admissions_tble$admit_day <- day(admissions_tble$admittime)
ggplot(admissions_tble, aes(admit_day)) + 
  geom_freqpoly()
```
Understandably, there are no dates at 0 or at 32. Most dates seem to occur 
in the same general frequency as the others, besides 14. 

#### Admission Week Day
```{r}
admissions_tble$admit_wd <- wday(admissions_tble$admittime, label = TRUE)
ggplot(admissions_tble, aes(admit_wd)) + 
  geom_bar(fill = "lightblue", color = "darkblue")
```
The days of the week seem to occur in the same frequencies. 

#### Admission Hour
```{r}
admissions_tble$admit_hour <- hour(admissions_tble$admittime)
ggplot(admissions_tble, aes(admit_hour)) + 
  geom_histogram(color = "darkblue", fill = "lightblue")
```
Admission hours are distributed in such a way that random hours throughout the 
day are missing. 

#### Admission Minute
```{r}
admissions_tble$admit_min <- minute(admissions_tble$admittime)
ggplot(admissions_tble, aes(admit_min)) + 
  geom_histogram(color = "darkblue", fill = "lightblue")
```
Most of the minutes occur at the same frequency, besides an unusually high 
number of admits on the hour, and what seems to be 15 the 15 minute intervals
(15, 30, and 45).

#### Length of Hospital Stay
```{r}
ggplot(icustays_tble, aes(los)) + 
  geom_histogram(color = "darkblue", fill = "lightblue", binwidth = 1)
```
It's strange that there are patients who stayed in the ICU for over 3 months. 
Majority of patients stay in the ICU for about 1 or 2 days. The bin width was set
to 1 because `los`, the length of stay variable, is measured in fractional days.

## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/hosp/patients/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/core/patients.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).
```{r}
setwd("~/mimic/core")
patients_tble <- as_tibble(read_csv("patients.csv.gz", show_col_types = FALSE)) %>%
  inner_join(icustays_tble, by = "subject_id")
```

2. Summarize variables `gender` and `anchor_age`, and explain any patterns you see.
```{r}
patients_tble %>% ggplot(aes(gender)) +
  geom_bar(fill = "lightblue", color = "darkblue")
patients_tble %>% ggplot(aes(anchor_age)) + 
  geom_histogram(fill = "lightblue", color = "darkblue", binwidth = 5)
```
We can see that there are more females than males, and that `anchor_age` is very
left skewed. 

## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/hosp/labevents.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```
`d_labitems.csv.gz` is the dictionary of lab measurements. 
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/hosp/d_labitems.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Find how many rows are in `labevents.csv.gz`.

2. We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), and glucose (50931). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`. 

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` option in the `read_csv` function to reduce the memory burden. It took my computer 5-10 minutes to ingest this file. If your computer really has trouble importing `labevents.csv.gz`, you can import from the reduced data file `labevents_filtered_itemid.csv.gz`.

3. Further restrict `labevents_tble` to the first lab measurement during the ICU stay. 

4. Summarize the lab measurements by appropriate numerics and graphics. 

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/chartevents.csv.gz"), 
    " | head"), 
  intern = TRUE
)
```
`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/d_items.csv.gz"), 
    " | head"), 
  intern = TRUE
)
```

1. We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

    Hint: `chartevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` option in the `read_csv` function to reduce the memory burden. It took my computer >15 minutes to ingest this file. If your computer really has trouble importing `chartevents.csv.gz`, you can import from the reduced data file `chartevents_filtered_itemid.csv.gz`.

2. Further restrict `chartevents_tble` to the first vital measurement during the ICU stay. 

3. Summarize these vital measurements by appropriate numerics and graphics. 

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are the first ICU stay of each unique adult (age at admission > 18) and columns contain at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- all variables in `patients.csv.gz`  
- first lab measurements during ICU stay  
- first vital measurements during ICU stay
- an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)

## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

- `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)

- `thirty_day_mort` vs first lab measurements

- `thirty_day_mort` vs first vital measurements

- `thirty_day_mort` vs first ICU unit

